Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-04-16 VegOzone.tex
Comments and suggestions from Lisa Sec 2-3

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on the 2018 growing season which was subject to an extended heatwave and its subsequent growing season of 2019. The NIBIO operated Environment Centre Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley serves as the main reference. Throughout the paper, we will use Svanhovd synonymously for the atmospheric monitoring site at Svanvik/Pasvik. The available meteorological monitoring data are described in Section~\ref{subsec:atmo_svanvik}. Other data used for cross-evaluation of the ozone monitoring data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use these to evaluate reanalysis products for ground-level ozone (Section~\ref{subsubsec:ozone_rea}).

\subsection{NIBIO Environment Centre Svanhovd and atmospheric monitoring}
\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. The locations of the atmospheric monitoring site and the ozone garden are marked in aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik and the border to Russia, the Pasvik river, lies only a few hundred meters to the east.\\

@@

An ozone garden was installed in the course of our project at the NIBIO Environment Centre Svanhovd and was likely the northern-most ozone garden to date. An ozone garden consists of selected plant species that are more sensitive to ozone and display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

To relate observed damage on ozone sensitive plants to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19 in addition to existing monitors of atmospheric key variables. All variables were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. Our ozone monitoring data has also been added to the EBAS database operated by \citet{NILU_EBAS} and is thus openly accessible.
Long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.
All relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data were recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

 As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.
Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in spring (May) and summer (July) 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

@@

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we derive a ground-level ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated continuously by NILU at Svanhovd, but no ozone measurements have been conducted until the start of our project. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we will cross-calibrate our data with long-term ozone observations at other sites in northern Fennoscandia. Concerning ozone, available observational data are sparse in this region. An overview of the geographical locations of conducted ozone monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data are described in Section~\ref{subsubsec:ebas}. 
As an additional complement to observational data which are not available at all times and spatially sparse, we also evaluate various ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{station_map_fennoscandia}
  \includegraphics[width=8.3cm]{fig03}
  \caption{Cap of the North. Locations of past and present ozone observation sites in northern Fennoscandia were used in this study. For more details see Tab.~\ref{tab:ebas_obs}. The same color coding for each station as displayed herein will be used in all the following figures.}
  \label{fig:station_map_fennoscandia}
\end{figure}

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview of the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, in close vicinity, the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above treeline (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass (c.c. Fig.\ref{fig:svanhovd_research_station}).

Data from before the 1990s are available solely from Svanvik and Karasjok. These data, however, do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably leads to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a breaking point in the combined data series, denoted as Jergul/Karasjok in the following, which would be a problem if this was the sole data to compare with. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are shown for completeness. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas indicate periods with insufficient quality control as mentioned above.

@@

\begin{figure*}[t]
  %\includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \includegraphics[width=12cm]{fig04}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}) (1986--2019). Data are available from EBAS. The hatched areas indicate periods with insufficient quality control according to \citet{NILU2003}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone, the Monitoring Atmospheric Composition and Climate (MACC) and the newer Copernicus Atmosphere Monitoring Service reanalysis (CAMSRA). From Copernicus member institutions a CAMS Regional Air Quality (CAMSRAQ) reanalysis ensemble is available for Europe. The CAMSRA and CAMSRAQ are available in near real time, but only the former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, the latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite-derived tropospheric column ozone, while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone near-surface station networks are only assimilated in the CAMSRAQ reanalysis ensemble. The quality concerning observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. In summary, the results prevented us from using any of the global reanalysis data.

@@

\section{Statistical analysis of environmental key variables}
\label{sec:stats}
In this section, we will evaluate atmospheric conditions in the growing seasons of 2018/19 by statistical means. We will derive climatologies for temperature, precipitation, global irradiance, and ozone for Svanvik (Section~\ref{subsec:climatologies}). The ozone measurements will be put into a climatological as well as regional context. We will look at anomalies of the atmospheric key variable mentioned above (Section~\ref{subsec:anomalies}).

@@

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the two dimensional density distributions for the combined northern Fennoscandia data (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} are displayed together with $1 \sigma$ uncertainties and SE, respectively. Splines have been fitted through the data for easier inter-comparison with reanalysis data.
From the northern Fennoscandia climatology, one can deduce several features. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. A sharp increase starts one month ahead of the estimated begin of greening $G_\mathrm{begin}$ (see Section~\ref{subsec:do3se_parameters} for details) and is in coincidence with rising temperatures and global irradiance. We find day of the year (\unit{doy}) $\left<G_\mathrm{begin}\right> = (150 \pm 14)$ for Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} is reached in late April ($100-120\,\unit{doy}$). The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$. The minimum occurs at $25\,\unit{ppb}$ in the beginning of September ($240-250\,\unit{doy}$).

The density distribution for Svanvik and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display a similar pattern as the northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier (at the beginning of April). As Svanvik is located at a lower altitude than the other sites, the earlier decline of the ozone peak may point to a stronger influence of vegetation on the removal of ozone. \chem{CO_2} uptake in coniferous trees can occur as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during nighttime.

@@

Finally, we shall have a look at the climatologies derived from observational ozone data and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed ground-level ozone well. The MACC reanalysis (Fig.~\ref{fig:ozone_climatologies}a) is in general too low ($(9\pm 7)\,\unit{ppb}$) and displays an erroneous seasonal cycle. Ozone concentrations peak already in March followed by the second peak in July. Also CAMSRA matches the observed ozone climatology poorly (Fig.~\ref{fig:ozone_climatologies}b). Despite displaying only one ozone peak, it still does not reproduce the seasonality in northern Fennoscandia well. The CAMSRA spring peak lags behind observations by $1\,\unit{month}$ and is $5\,\unit{ppb}$ too low, whereas the minimum occurs in January compared to August/September. The annual amplitude ($(26\pm 1)\,\unit{ppb}$) is larger than in the climatology derived from observations ($19\,\unit{ppb}$). The modeled \chem{[O_3]} at Svanvik are slightly higher than at Pallas which might be due to the height at which surface ozone has been evaluated in the model. In contrast, the ensemble mean of the CAMSRAQ reproduces the seasonal cycle in northern Fennoscandia quite well (Fig.~\ref{fig:ozone_climatologies}c), but shows on average a slight low-bias ($(2.8\pm 0.5)\,\unit{ppb}$).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatologies_portrait}
  \includegraphics[width=8.3cm]{fig07}
  \caption{Comparison of climatologies derived from observational data (EBAS) with various ozone reanalysis products: (a) MACC reanalysis; (b) CAMSRA; (c) CAMSRAQ reanalysis ensemble mean. As the representative for the sites, the nearest neighboring grid point has been chosen. The global products (a), (b) do not reproduce observed ground-level ozone well. The MACC reanalysis (a) is in general too low and displays an erroneous seasonal cycle. The CAMSRA (b) matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of CAMSRAQ reanalysis ensemble reproduces the seasonal cycle in northern Fennoscandia, though the modeled ozone concentrations are slightly too low.}
  \label{fig:ozone_climatologies}
\end{figure}

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The model enhancements from MACC reanalysis to CAMSRA have been reported and discussed by \citet{ACP:Inness2019} on a global level. Amongst others, assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme has lead to the enhanced performance of the latter. The higher spatial resolution of the regional air quality models used in the CAMSRAQ ensemble compared to the global reanalyses captures small scale depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models concerning daily and seasonal cycles in ozone. The latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results is the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough \citep{ACP:Andersson2017}. \citet{ACPD:Barten2020} reported pronounced biases for CAMSRA in the high Arctic (e.g., Summit, Greenland) and attribute this to dry deposition to the ocean.

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
Using these established climatologies, we look at the anomalies in 2018/19 and discuss the differences between the two years.

@@

In the following, we will use the interchangeable terms anomalies and residuals to distinguish between a climatological (anomalies) and a statistical (residuals) point of view.
All the above factors make it necessary to look not only at anomalies but statistically assess ozone concentration residuals at Svanvik as well as at Esrange and Pallas in comparison. Since the climatology of the latter is more reliable, we will start with ground-level ozone concentration residuals \chem{\Delta[O_3]} at Esrange and Pallas in 2018. As data for 2019 has not yet been available from EBAS (last access, February 2020), we refrain from looking into these.

>>>>

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on comparing the anomalous warm, bright and dry 2018 growing season with the more climatically average 2019 season to understand the role these environmental conditions may have played in influencing the sensitivity of vegetation to ozone damage. The NIBIO operated Environment Centre Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley serves as the main reference for conditions in northern Fennoscandia. Throughout the paper, we will use Svanhovd synonymously for the atmospheric monitoring site at Svanvik/Pasvik. The available meteorological monitoring data are described in Section~\ref{subsec:atmo_svanvik}. Other data used for cross-evaluation of the ozone monitoring data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We, in particular, compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use these to evaluate reanalysis products for ground-level ozone (Section~\ref{subsubsec:ozone_rea}).

\subsection{NIBIO Environment Centre Svanhovd and atmospheric monitoring}
\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. The locations of the atmospheric monitoring site and the ozone garden are marked in aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik, where the Pasvik river, which is the border to Russia, lies only a few hundred meters to the east.\\

@@

An ozone garden was installed over the course of our project at the NIBIO Environment Centre Svanhovd and was likely the northern-most ozone garden to date {\bf TODO: Reference?}. An ozone garden consists of selected plant species that are more sensitive to ozone and display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

To relate observed damage on ozone sensitive plants to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19 in addition to existing monitors of atmospheric key variables. All variables were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. Our ozone monitoring data has also been added to the EBAS database operated by \citet{NILU_EBAS} and is thus openly accessible.
Long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.
All relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations \chem{[O_3]} measured at $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data were recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

 As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. We present a reconstruction concerning the missing data in Section~\ref{subsec:ozone_reco}. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.
Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in spring (May) and summer (July) 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

@@

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the higher \chem{[O_3]} in 2018 in relation to regional averages and trends, we derive a ground-level ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated continuously by NILU at Svanhovd, but no ozone measurements have been conducted until the start of our project. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we cross-calibrate our data with long-term ozone observations at other sites in northern Fennoscandia. Concerning ozone, available observational data are sparse in this region. An overview of the geographical locations of conducted ozone monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data are described in Section~\ref{subsubsec:ebas}. 
As an additional complement to observational data which are not available at all times and spatially sparse, we also evaluate \chem{[O_3]} derived from a variety of reanalysis products (Section~\ref{subsubsec:ozone_rea}).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{station_map_fennoscandia}
  \includegraphics[width=8.3cm]{fig03}
  \caption{Cap of the North. Locations of past and present (end of 2018) ozone observation sites in northern Fennoscandia were used in this study. For more details see Tab.~\ref{tab:ebas_obs}. The same color coding for each station as displayed herein will be used in all the following figures.}
  \label{fig:station_map_fennoscandia}
\end{figure}

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview of the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in relatively higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at higher resolution the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above treeline (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the WMO standards {\bf TODO: Reference \url{https://library.wmo.int/index.php?id=12407&lvl=notice_display}} (Fig.\ref{fig:svanhovd_research_station}).

Data from before the 1990s are available solely from Svanvik and Karasjok. These data, however, do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably leads to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a breaking point in the combined data series, denoted as Jergul/Karasjok in the following, which would be particularly problematic if these were the sole data for comparison. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are shown for completeness. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas indicate periods with insufficient quality control as mentioned above.

@@

\begin{figure*}[t]
  %\includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \includegraphics[width=12cm]{fig04}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}) (1986--2019). Data from EBAS until December 2018. The hatched areas indicate periods with insufficient quality control according to \citet{NILU2003}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.

There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone, the Monitoring Atmospheric Composition and Climate (MACC) and the newer Copernicus Atmosphere Monitoring Service reanalysis (CAMSRA) \citep{ACP:Inness2013, ACP:Inness2019}. The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75\,\unit{^\circ}\times0.75\,\unit{^\circ}$ or roughly $29.3\,\unit{km}\times 83.4\unit{km}$ at the location of Svanvik. From Copernicus member institutions, higher resolution suface ozone ranalysis data are available for Europe (CAMS Regional Air Quality (CAMSRAQ) reanalysis ensemble). The CAMSRAQ is based on nine European state-of-the-art numerical air quality models. The ensemble mean is at fairly high spatial and temporal resolution compared to the global reanalysis: $0.1\,\unit{^\circ}\times 0.1\,\unit{^\circ}$ or roughly $3.9\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The periodes covered differ and no data is available before the turn of the millennium. The CAMSRA and CAMSRAQ are available in near real time, but only the former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, the latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite-derived tropospheric column ozone, while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone near-surface station networks are only assimilated in the CAMSRAQ reanalysis ensemble. The quality concerning observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. Due to the found substantial discrepancies between reanalysis data and site observations in northern Fennoscandia, we refrain from using any of the global reanalysis data.

@@

\section{Statistical analysis of environmental key variables}
\label{sec:stats}
To improve our understanding and determine the risk of vegetation from high \chem{[O_3]} in subarctic biomes in a changing climate, we first assess the climate conditions and evaluate the growing seasons 2018/19 concerning anomalies from multi-annual means (climatologies). We derive climatologies for temperature, precipitation, global irradiance, and ozone for Svanvik. The ozone measurements will be put into a climatological as well as regional context.

@@

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the two dimensional density distributions of \chem{[O_3]} for the combined northern Fennoscandia data (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} are displayed together with $1 \sigma$ uncertainties and SE, respectively. Splines have been fitted through the data for easier inter-comparison with reanalysis data.
From the northern Fennoscandia climatology, one can deduce several features. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. A sharp increase starts one month ahead of the estimated begin of greening $G_\mathrm{begin}$ (see Section~\ref{subsec:do3se_parameters} for details) and is in coincident with rising temperatures and global irradiance. We find day of the year (\unit{doy}) $\left<G_\mathrm{begin}\right> = (150 \pm 14)$ for Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates to the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} is reached in late April ($100-120\,\unit{doy}$). The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$. The minimum occurs at $25\,\unit{ppb}$ in the beginning of September ($240-250\,\unit{doy}$).

The density distribution for Svanvik and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display a similar pattern as the northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier (at the beginning of April). As Svanvik is located at a lower altitude than the other sites, the earlier decline of the ozone peak may point to a stronger influence of vegetation on the removal of ozone. \chem{CO_2} uptake by coniferous trees can occur as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during nighttime.

@@

Finally, we shall have a look at the climatologies derived from observational ozone data and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). To represent the sites, the nearest neighboring grid point of the \chem{[O_3]} reanalysis products is chosen. The global products do not reproduce observed ground-level \chem{[O_3]} well. This is important for European wide and global risk assessment that rely on these data. The MACC reanalysis (Fig.~\ref{fig:ozone_climatologies}a) is in general too low ($(9\pm 7)\,\unit{ppb}$) and displays an erroneous seasonal cycle. Ozone concentrations peak already in March followed by the second peak in July. Also CAMSRA matches the observed ozone climatology poorly (Fig.~\ref{fig:ozone_climatologies}b). Despite displaying only one ozone peak, it still does not reproduce the seasonality in northern Fennoscandia well. The CAMSRA spring peak lags behind observations by $1\,\unit{month}$ and is $5\,\unit{ppb}$ too low, whereas the minimum occurs in January compared to August/September. The annual amplitude ($(26\pm 1)\,\unit{ppb}$) is larger than in the climatology derived from observations ($19\,\unit{ppb}$). The modeled \chem{[O_3]} at Svanvik are slightly higher than at Pallas which might be due to the height at which surface ozone has been evaluated in the model. In contrast, the ensemble mean of the CAMSRAQ reproduces the seasonal cycle in northern Fennoscandia quite well (Fig.~\ref{fig:ozone_climatologies}c), but shows on average a slight low-bias ($(2.8\pm 0.5)\,\unit{ppb}$).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatologies_portrait}
  \includegraphics[width=8.3cm]{fig07}
  \caption{Comparison of climatologies derived from observational data (gray band) with three ozone reanalysis products: (a) MACC reanalysis; (b) CAMSRA; (c) CAMSRAQ reanalysis ensemble mean. The sites, represented by the nearest neighboring grid point of the \chem{[O_3]} reanalysis products, are displayed in different markers and colors. The global products (a), (b) do not reproduce observed ground-level ozone well. The MACC reanalysis (a) is in general too low and displays an erroneous seasonal cycle. The CAMSRA (b) matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of CAMSRAQ reanalysis ensemble reproduces the seasonal cycle in northern Fennoscandia, though the modeled ozone concentrations are slightly too low.}
  \label{fig:ozone_climatologies}
\end{figure}

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The model enhancements from MACC reanalysis to CAMSRA have been reported and discussed by \citet{ACP:Inness2019} on a global level. Amongst others, assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme has lead to the enhanced performance of the latter, but a considerable bias still remains \citep{GMD:Huijnen2020}. In particular, \citet{ACPD:Barten2020} reported a pronounced bias for CAMSRA in the high Arctic (e.g., Summit, Greenland) and attribute this bias to dry deposition to the ocean. The higher spatial resolution of the regional air quality models used in the CAMSRAQ ensemble compared to the global reanalyses captures small scale depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models concerning daily and seasonal cycles in ozone. The latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results is the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough \citep{ACP:Andersson2017}.

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
Using these established climatologies, we look at the anomalies in 2018/19 and discuss the differences between the two years. As weather extremes like in 2018 are likely to become more frequent in occurrence with climate change, this will help improving our understanding of the role which these climate variables have on determining the sensitivity of subarctic vegetation to \chem{[O3]}.

@@

In the following, we will use the interchangeable terms anomalies and residuals to distinguish between a climatological (anomalies) and a statistical (residuals) point of view.
All the above factors make it necessary to look not only at anomalies but statistically also to assess ozone concentration residuals at Svanvik as well as at Esrange and Pallas in comparison. Since the climatology of the latter is more reliable, we will start with ground-level ozone concentration residuals \chem{\Delta[O_3]} at Esrange and Pallas in 2018. As data for 2019 has not yet been available from EBAS (last access, February 2020), we refrain from looking into these.

In Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals}, the probability density functions of daily mean ozone concentration residuals for Esrange and Pallas are shown. Both stations display a clear, but not significant, deviation from the climatological mean. On average, \chem{[O_3]} had been elevated by $(1.9\pm 5.7)\,\unit{ppb}$ and $(1.9\pm 5.4)\,\unit{ppb}$, respectively. Both distributions are slightly skewed towards positive deviations with $\frac{1}{4}$ of days (3. quantile, q3) falling above $5.8\,\unit{ppb}$ and $5.6\,\unit{ppb}$, respectively.
